---
title: 'Analysis of Microarray Data 1 (E-MTAB-2853)'
output:
  pdf_document:
    dev: png
    toc: true
  html_document:
    df_print: paged
    toc: true
header-includes: \setlength{\emergencystretch}{1em}
---

```{r global_options, include=FALSE, message=F, echo=F}
knitr::opts_chunk$set(fig.align = 'center')
knitr::opts_chunk$set(out.width = '90%')
knitr::opts_chunk$set(fig.width=12, fig.height=8)

library("ArrayExpress")
library("arrayQualityMetrics")
library("limma")
library("siggenes")
library("affy")
library("htmltools")
library("dplyr")
library("kableExtra")
```

# 1. Data

## General info
Microarray type: RNA samples prepared for hybridization to Affymetrix microarrays (E-MTAB-2853).

## Load sdrf file
For our analysis, we will use data from the seedlings induced for four hours with dexamethansone or a mock solution.
```{r}
sdrf <- read.delim("./Data/Microarray/E-MTAB-2853.sdrf.txt")
# select only the seedlings induced with dexamethansone or a mock solution
sdrf <- as.data.frame(sdrf[c(5,6,7,8,9,10,11,12),])
rownames(sdrf) <- sdrf$Array.Data.File
# correct value from "Factor.Value..time" column of row "mz_210906_mz_D_1_B.CEL"
# 4 -> 1
sdrf$Factor.Value..time.[2] <- 1
sdrf[,c("Source.Name","Array.Design.REF", "Factor.Value..compound.",
        "Factor.Value..time.")]
```

## Load in the data
```{r}
# read in microarray data
exonCELs <- list.celfiles("./Data/Microarray/")
data.raw <- ReadAffy(filenames = exonCELs, celfile.path = "./Data/Microarray/", 
                     phenoData = sdrf)
# rename column names of phenoData
pData(data.raw) <- rename(pData(data.raw), "treatment"="Factor.Value..compound.", 
                          "time"="Factor.Value..time.")
# rename treatment values of phenoData
pData(data.raw)$treatment <- pData(data.raw) %>%
  select(treatment) %>%
  mutate(
    treatment = recode(treatment,"Dexamethasone"="DEX", "mock solution"="mock")
  )
pData(data.raw)$treatment <- pData(data.raw)$treatment[[1]]
```

## Quality Control on raw data

arrayQualityMetrics (open "index.html" file for a full overview of the output)
```{r, include = FALSE, message=F}
arrayQualityMetrics(data.raw, outdir="./Data/Microarray/raw", force=T)
arrayQualityMetrics(data.raw, outdir="./Data/Microarray/rawlog", force=T, 
                    do.logtransform=T)
```

# 2. Preprocessing
The preprocessing includes background correction, between array intensity adjustment (normalization) and probeset summarization.
```{r, include = FALSE, message=F}
data.rma<- affy::rma(data.raw,background=T)
arrayQualityMetrics(data.rma,outdir="./Data/Microarray/rma_processed",force=T)  			#RMA produces log-transformed data
```

```{r}
# Before preprocessing
dim(exprs(data.raw))
# After preprocessing
dim(exprs(data.rma))
```

Check data distribution
```{r}
boxplot(exprs(data.rma), names=data.rma$Source.Name)
```
Phenotypic data and the intensity values
```{r}
pData(data.rma)
head(exprs(data.rma))
```

# 3. Differential Analysis

## Design Matrix and Contrast

Set the effect of interest (mock vs Dexamethasone treatment) as a factor, as well as the blocking time effect.
```{r}
pd <- pData(data.rma)
pd$treatment <- pd$treatment
treatment <- pd$treatment
pd$time <- as.factor(pd$time)
time <- pd$time
```

Define design matrix
```{r}
design <- model.matrix(~treatment+time)
colnames(design) <- c("Intercept","mock.vs.DEX","time_4h")
design
contrast.matrix <- makeContrasts(TvsN="mock.vs.DEX",
                                 levels=design)
```

## Fitting the model
```{r}
fit <- lmFit(data.rma, design)
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)
```

## Getting top genes
```{r}
summary_fit2 <- summary(decideTests(fit2,method="separate",adjust.method="BH",
                                    p.value=0.05,lfc=0))
t(summary_fit2)
LIMMAout <- topTable(fit2,adjust="BH",number=nrow(exprs(data.rma)))
```

# 4. Identification of Differential Expression Genes (DEGs)

## Annotate gene ID to respective probes
```{r}
annotation_MA = read.table(file="./Data/Microarray/A-AFFY-2.adf.txt", 
                           header=TRUE, sep="\t", fill=TRUE, skip=12)
annotation_MA <- 
  annotation_MA[sort(annotation_MA$Composite.Element.Name,index.return=T)$ix,]
# Check if all probes are present in both sets
table(annotation_MA$Composite.Element.Name==sort(rownames(LIMMAout)))
# Sort LIMMA output alphabetically on probe name
LIMMAout_sorted <- LIMMAout[sort(rownames(LIMMAout),index.return=T)$ix,]
# Add gene names to LIMMA output
LIMMAout_sorted$gene_ID <- toupper(annotation_MA$Composite.Element.Database.Entry.tair.)
LIMMAout_annot <- 
  LIMMAout_sorted[sort(LIMMAout_sorted$adj.P.Val,index.return=T)$ix,]

# Filter genes that underexpressed after DEX treatment
LIMMAout_underexpressed <- LIMMAout_annot %>% dplyr::select(everything()) %>%
  filter(adj.P.Val < 0.05 & logFC > 1) %>%
  arrange(desc(logFC))

# Filter genes that overexpressed after DEX treatment
LIMMAout_overexpressed <- LIMMAout_annot %>% dplyr::select(everything()) %>%
  filter(adj.P.Val < 0.05 & logFC < -1) %>%
  arrange(logFC)

# Manually added descriptions for 10 top significantly overexpressed and underexpressed genes
# There were only 11 overexpressed gene so only selected 11
LIMMAout_top15 <- rbind(LIMMAout_underexpressed[1:15,], LIMMAout_overexpressed[1:11,])
```

## Annotate genes to respective gene IDs
```{r message=F}
library("org.At.tair.db")
annotated_geneID <- AnnotationDbi::select(org.At.tair.db, 
                keys=LIMMAout_top15$gene_ID, "SYMBOL", "GENENAME", keytype = 'TAIR')

LIMMAout_top15$gene_symbol <- NA

for (tair in LIMMAout_top15$gene_ID){
  LIMMAout_top15[LIMMAout_top15$gene_ID==tair,]$gene_symbol <- 
    paste(annotated_geneID[grepl(tair, annotated_geneID$TAIR),]$SYMBOL, collapse=",")
}

LIMMAout_top15 <- LIMMAout_top15[,c(7,8,1,4,5)]
LIMMAout_top15 %>%
  kbl(caption = "Top-15 Significant Genes", longtable=T, booktabs = T, row.names=F) %>%
kable_styling(latex_options="striped", position="center", full_width=F)
```

## Plotting results
```{r}
volcanoplot(fit2, col=as.factor(LIMMAout$adj.P.Val < 0.05), 
            style="p-value", highlight = 10)
```

```{r}
limma::plotMA(fit2)
```

```{r}
plot(LIMMAout$AveExpr, LIMMAout$logFC,
  col=as.factor(LIMMAout$adj.P.Val < 0.05), pch=20, cex=0.50,
  xlab="Average Intensities", ylab="logFC")
```

# 5. Gene Set Analysis

```{r}
LIMMAout_filtered <- LIMMAout_annot[(LIMMAout_annot$adj.P.Val<0.09),]
# 304 significant genes are filtered
length(LIMMAout_filtered$tair)
# save significant gene identifiers of significant genes (adj.P.Val<0.09)
write.table(LIMMAout_filtered$tair,"Microarray_SignGenes.csv",sep=",")
```

## Compute GO enrichment analysis in PLAZA 5.0 Workbench

```{r}
GO_enrich_res <- read.table(
  file="./Data/Microarray/GO_enrichment_analysis_MA1.csv", sep=",", header = TRUE)
GO_enrich_res$FDR <- p.adjust(GO_enrich_res$P.Value, method="BH")
GO_enrich_res <- GO_enrich_res[,c(1,6,2,3,4,8,5)]
colnames(GO_enrich_res) <- c("GOTerm", "Description", "Ontology", "logFC", 
                             "P.Value", "FDR", "Subset ratio(%)")
GO_enrich_res_BP <- GO_enrich_res[GO_enrich_res$Ontology=="BP",]
rownames(GO_enrich_res_BP) <- c(1:length(GO_enrich_res_BP$GOTerm))

# Filter GO terms associated to biological processes
GO_enrich_res_BP %>%
  kbl(caption = "GO Enrichment Result (PLAZA 5.0 Workbench)", 
longtable=T, booktabs = T) %>%
  kable_styling(latex_options="striped", position="center", full_width=F)
```
# 6. Save results for comparison

```{r}
# Result from limma analysis
write.table(LIMMAout_annot, sep="\t", file="Microarray_DE_results.txt")
# Result from Gene Set Analysis
write.table(GO_enrich_res_BP, sep="\t", file="Microarray_GSA_results.txt")
```